\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2018-01-22 eg6-letters/NIM/EG6_RTPC_NIM.tex

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nucleus) necessitates an upgrade of the photon detection system.  
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 experiment 
in 2005~\cite{Girod:2007aa}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the CLAS detector and the IC from the 
large flux of low energy M{\o}ller electrons, a 5~T solenoid magnet was placed 
around the target to shield the detectors.  To detect recoiling $\alpha$ 
particles from the coherent DVCS on Helium, a new radial time projection 
chamber (RTPC) was developed to track low energy nuclear fragments. The 
solenoid field was used to bend tracks and measure momentum of particles in the 
RTPC. The CLAS detector supplemented with both IC and RTPC was used during a 
three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 
2005~\cite{Fenker:2008zz}. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. 
Moreover, the data aquisition electronic was improved to increase the event 
readout rate. The enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} of 
this paper. The data acquisition system is described in section \ref{sec_readout}, 
the calibration methods in section \ref{sec_calib} and the tracking algorithm in section
\ref{sec_tracking}. Finally, the overall performances
of the RTPC are described in section \ref{sec_perfor}.

@@

We detail here the different regions shown in Figure~\ref{fig:RTPC_1_4} 
starting from the beam line towards larger radius:\\
\begin{itemize}
  \item The 6~atm $^4$He target extends along the beamline forming the detector 
     central axis. It is a 6~mm diameter Kapton straw with a 27~\textmu{}m wall 
     of 292~mm length such that its entrance and exit 15~\textmu{}m aluminum windows 
     are placed outside of the detector volume. The detector and the target are 
     placed in the center of the solenoid, 64~cm upstream of the CLAS center.
   \item The first gas gap covers the radial range from 3~mm to 20~mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 \textmu{}m thick window made of grounded aluminized Mylar.
   \item The second gas gap region extends between 20~mm and 30~mm and is 
      filled with the gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether 
      (DME).  This region is surrounded by a 4~\textmu{}m thick window made of 
      aluminized Mylar set at $-4260$~V to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and 
      extends from the cathode to the first GEM, 60~mm away from the beam axis.  
      The average electric field in this region is perpendicular to the beam 
      and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66~mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600~V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charges 
      after they have been multiplied by the GEMs. Pre-amplifiers are plugged 
      directly on its outer side and transmit the signal to the data 
      acquisition electronics.
\end{itemize}

@@

The main structural improvement compared to the 2005 RTPC~\cite{Fenker:2008zz} 
was to obtain a better support structure for the GEM foils. In the previous version,
the RTPC was essencially composed of two independent half-cylinder separated
by their own structures. In this design, the installation of the GEMs was not very practical
and wrinkles were visibles on the GEM surfaces. To allow for a better installation 
and, at the same time, keep the mechanical structure out of the drift volume, 
the present design is based on self supporting GEM cylinder. We 
realized these cylinders by using fiber glass rings glued to 
each ends of the GEM foils in order to 
instal them independently in the RTPC after gluing and soldering operations. The 
rigidity of the GEM foils was enough for the structure to be self-supporting 
and only the upstream end of the cylinder was fixed to the main mechanical 
support structure. This design only left a light fiberglass ring in the 
downstream end, reducing to a minimum secondary interactions.

@@

The readout system was an upgraded version of the original BoNuS RTPC system~\cite{Fenker:2008zz}, 
based ALICE-TPC front end electronic boards (FECs)~\cite{ALICE-FEE}. 
Each FEC hosted 128 channels, providing amplification and digitization of the 
input signals. For each event, 100 samples/channel were digitized by the ALTRO 
ASIC~\cite{EsteveBosch:2003bj}, operated at 10 MHz sampling frequency. In order 
to reduce the data size, the system was operated in zero-suppression mode, 
keeping data from $N_{PRE}=3$ samples before threshold crossing on the rising 
edge to $N_{POST}=3$ samples after threshold crossing on the falling edge. 
The threshold level was set on each channel just above the noise level.

@@

The first step in the calibration is to fix the maximum drift time ($T_{max}$) that an
electron can arrive on the detection plane. This value is highly dependent on the gas 
mixture composition, the electric and the magnetic field experienced by the drift 
electrons. This time correspond to the electrons released close to 
the cathode, which take the most time to reach the readout pads as illustrated in 
Figure~\ref{fig:RTPC_signals}. Note that the minimum time is not calibrated, as we
use the trigger time is defined ($T_{min}=15$), and that the time unit is the ALTRO 
sampling time of 100~ns.

\begin{figure}[tb]
\centering
\includegraphics[{trim={0 12cm 0 0},clip, scale=0.4}]{RTPC_1_4.pdf}
\caption[]{A schematic drawing of a $^{4}$He track (in green) traversing the 
drift region, with the drift paths followed by the electrons (in black). } 
\label{fig:RTPC_signals}
\end{figure}

In order to experimentally determine the value of $T_{max}$, we analyse the
time profile of hits from identified good tracks, as presented in 
Figure~\ref{fig:TDC_profile}. We can clearly observe the expected dropping 
edge and define $T_{max}$, as the time at which the distribution passes 
below half the maximum number of hits in the histogram. This value was 
measured in bins along the 200~mm RTPC's 
length to take into account variations in the electric and magnetic field in 
the RTPC. We show this result for one experimental run in 
Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}.

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{hits_time_profile.pdf}
\caption{Time distribution of the hits in one z-bin along the RTPC in one 
experimental run. } \label{fig:TDC_profile}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{RunNumber_61452_TDCmax_Zslice.pdf}
\caption{Maximum time of collected hits as a function of the track
              position on the $z$-axis for one experimental run. } 
\label{fig:RunNumber_61551_TDCmax_Zslice}
\end{figure}

Figure~\ref{fig:Drift_run_number_1} shows the $T_{max/2}$ values for individual 
runs (approximately 2 hours long). We observe significant change of $T_{max}$
before and after run 61600, while variations within these periods are 
around 2\%. This is likely due to non-perfect experimental conditions, in 
particular possible contamination of our gas mixture. We indeed increase the gas flow 
around that time due to a small leak in the RTPC. While our 
gas system was kept slightly over atmospheric pressure to limit 
contamination from air or other external gases, it is likely that this leak was 
the source of modification of the drift velocity.

\begin{figure*}[tb]
\centering
\includegraphics[width=15.5cm]{Drift_run_number.pdf}
\caption{$T_{max/2}$ versus the experimental run numbers.} 
\label{fig:Drift_run_number_1}
\end{figure*}

@@

Based on the kinematics of the electrons in the calibration data, we generated 
the Helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then,
we compared the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. To perform the drift path extraction, 
we made a first approximation assuming a linear dependence between the radius 
of emission of the charge and its time of detection, and then refined our 
result. Because of the magnetic field, the drift paths in the RTPC are not linear 
but the curvature is small and the process converged already on the 
second iteration.

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.46]{FitResult_p2_11.pdf}
\caption{$\Delta \phi$ versus T distribution for tracks
from one bin in longitudinal position along the RTPC. The black line represents 
the final drift paths in this bin.}
\label{fig:DELTA_PHI_TDC}
\end{figure}

>>>>

For certain experiments the base CLAS system was complemented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nucleus) necessitates an upgrade of the photon detection system.  
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 experiment 
in 2005~\cite{Girod:2007aa}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the CLAS detector and the IC from the 
large flux of the low energy M{\o}ller electrons, a 5~T solenoid magnet was 
placed around the target to shield the detectors.  To detect recoiling $\alpha$ 
particles from the coherent DVCS on Helium, a new radial time projection 
chamber (RTPC) was developed to track low energy nuclear fragments. The 
solenoid field was used to bend tracks and measure momentum of particles in the 
RTPC. The CLAS detector supplemented with both IC and RTPC was used in 2009 during a 
three months experimental run~\cite{proposal1,proposal2}
with a longitudinally polarized, 130~nA and 6.064 GeV electron beam
incident on a gaseous $^{4}$He target.

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 
2005~\cite{Fenker:2008zz}. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. 
Moreover, the data aquisition electronic was improved to increase the event 
readout rate. The enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} of 
this article. The data acquisition system is described in section \ref{sec_readout}, 
the calibration methods in section \ref{sec_calib} and the tracking algorithm in section
\ref{sec_tracking}. Finally, the overall performances
of the RTPC are described in section \ref{sec_perfor}.

@@

We detail here the different regions shown in Figure~\ref{fig:RTPC_1_4} 
starting from the beam line towards larger radius:\\
\begin{itemize}
  \item The 6~atm $^4$He target extends along the beamline forming the detector 
     central axis. It is a 6~mm diameter Kapton straw with a 27~\textmu{}m wall 
     of 292~mm length such that its entrance and exit 15~\textmu{}m aluminum windows 
     are placed outside of the detector volume. The detector and the target are 
     placed in the center of the solenoid, 64~cm upstream of the CLAS center.
   \item The first gas gap covers the radial range from 3~mm to 20~mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 \textmu{}m thick window made of grounded aluminized Mylar.
   \item The second gas gap region extends between 20~mm and 30~mm and is 
      filled with the gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether 
      (DME).  This region is surrounded by a 4~\textmu{}m thick window made of 
      aluminized Mylar set at $-4260$~V to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and 
      extends from the cathode to the first GEM, 60~mm away from the beam axis.  
      The average electric field in this region is perpendicular to the beam 
      and average around 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66~mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600~V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charges 
      after they have been multiplied by the GEMs. Pre-amplifiers are plugged 
      directly on its outer side and transmit the signal to the data 
      acquisition electronics.
\end{itemize}

@@

The main structural improvement compared to the 2005 RTPC~\cite{Fenker:2008zz} 
was to obtain a better support structure for the GEM foils. In the previous version,
the RTPC was essentially composed of two independent half-cylinder separated
by their own structures. In this design, the installation of the GEMs was not 
very practical and wrinkles were visible on the GEM surfaces. To allow for a 
better installation and, at the same time, keep the mechanical structure out of 
the drift volume, the present design is based on self supporting GEM cylinder.  
We realized these cylinders by using fiber glass rings glued to each end of the 
GEM foils in order to install them independently in the RTPC after gluing and 
soldering operations. The rigidity of the GEM foils was enough for the 
structure to be self-supporting and only the upstream end of the cylinder was 
fixed to the mechanical support structure. This design only left a light 
fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

The readout system was an upgraded version of the original BoNuS RTPC system~\cite{Fenker:2008zz}, 
based ALICE-TPC front end electronic boards (FECs)~\cite{ALICE-FEE}. 
Each FEC hosted 128 channels, providing amplification and digitization of the 
input signals. For each event, 100 samples/channel were digitized by the ALTRO 
ASIC~\cite{EsteveBosch:2003bj}, operated at 10~MHz sampling frequency. In order 
to reduce the data size, the system was operated in zero-suppression mode, 
keeping data from $N_{PRE}=3$ samples before threshold crossing on the rising 
edge to $N_{POST}=3$ samples after threshold crossing on the falling edge. 
The threshold level was set on each channel just above the noise level.

@@

The first step in the calibration is to fix the maximum drift time ($T_{max}$) that an
electron can arrive on the detection plane. This value is highly dependent on the gas 
mixture composition, as well as the electric and magnetic field experienced by the drift 
electrons. The maximum time corresponds to electrons released close to 
the cathode, which are the furthest from the readout pads as illustrated in 
Figure~\ref{fig:RTPC_signals}. Note that the minimum time is not calibrated, as we
use the trigger time for it ($T_{min}=15$), and that the time unit is the ALTRO 
sampling time of 100~ns.

\begin{figure}[tb]
\centering
\includegraphics[{trim={0 12cm 0 0},clip, scale=0.4}]{RTPC_1_4.pdf}
\caption[]{A schematic drawing of a $^{4}$He track (in green) traversing the 
drift region, with the drift paths followed by the electrons (in blue). } 
\label{fig:RTPC_signals}
\end{figure}

In order to experimentally determine the value of $T_{max}$, we analyse the
time profile of hits from identified good tracks, as presented in 
Figure~\ref{fig:TDC_profile}. We can clearly observe the expected dropping 
edge and define $T_{max}$, as the time at which the distribution passes 
below half the maximum number of hits in the histogram. This value was 
measured in bins along the 200~mm RTPC's length to take into account 
variations in the electric and magnetic field. We show this result for one 
experimental run in Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}.

\begin{figure}[t]
\centering
\includegraphics[scale=0.42]{hits_time_profile.pdf}
\caption{Time distribution of the hits in one z-bin along the RTPC in one 
experimental run. } \label{fig:TDC_profile}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.42]{RunNumber_61452_TDCmax_Zslice.pdf}
\caption{Maximum time of collected hits as a function of the track
              position on the $z$-axis for one experimental run. } 
\label{fig:RunNumber_61551_TDCmax_Zslice}
\end{figure}

Figure~\ref{fig:Drift_run_number_1} shows the $T_{max}$ values for individual 
runs (approximately 2 hours long). We observe significant change of $T_{max}$
before and after run 61600, while variations within these periods are 
around 2\%. This is likely due to non-perfect experimental conditions, in 
particular possible contamination of our gas mixture. We indeed increase the gas flow 
around that time due to a small leak in the RTPC. While our 
gas system was kept slightly over atmospheric pressure to limit 
contamination from air or other external gases, it is likely that this leak was 
the source of modification of the drift velocity.

\begin{figure*}[t!]
\centering
\includegraphics[width=17.5cm]{Drift_run_number.pdf}
\caption{$T_{max}$ versus the experimental run numbers.} 
\label{fig:Drift_run_number_1}
\end{figure*}

@@

Based on the kinematics of the electrons in the calibration data, we generated 
the Helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then,
we compared the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. To perform the drift path extraction, 
we made a first approximation assuming a linear dependence between the radius 
of emission of the charge and its time of detection, and then refined our 
result accounting for the curvature of the drift path. The curvature was small 
enough, such that the process converged already on the second iteration.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
Figure~\ref{fig:Drift_run_number_1}). We found very similar drift paths for the 
two data sets and concluded that any changes in the system only significantly 
affected the drift speed.

\begin{figure}[t]
\centering
\includegraphics[scale=0.48]{FitResult_p2_11.pdf}
\caption{$\Delta \phi$ versus T distribution for tracks
from one bin in longitudinal position along the RTPC. The black line represents 
our firt of the drift path in this bin.}
\label{fig:DELTA_PHI_TDC}
\end{figure}
